Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-25 docs/manuscript.tex
continue edits

Nonetheless, \citeA{johansen2007paradoxical} demonstrated that the inverse base-rate effect can occur without the traditional predictive learning design.
In one of the conditions in their Experiment 3, the canonical inverse base-rate design (including the shared cue) was implemented in a list format.
In this format, the trial-by-trial presentation of training items was turned into a list of 12 items fitted on a single page.
Subsequently, participants made judgements about new cases on a separate page.
In this condition, participants still exhibited a rare preference on $BC$ trials.
In another condition of Experiment 3, participants received the information about outcome frequencies as a summary before testing.
After learning about feature-label information in this manner, participants did not show the IBRE but was matching the base rate.
These experiments give us evidence about another boundary condition for the IBRE - itemised rather than summerised presentation of training items.
% Need to elaborate why this is relevant.

Additionally, there are at least three studies which directly look at error-driven processes in the IBRE.
\citeA{don2019learned}, demonstrated that on $AC$ trials, people fixated on $C$ longer than on $A$ both pre-responding during stimulus presentation and post-responding during feedback.
This fixation bias increased with more training.
They also observe greater fixation on $C$ on $AC$ trials, relative to $B$ on $AB$ trials.
Furthermore, \citeA{wills2014attention} in an EEG study observed posterior selection negativity and concurrent frontal positivity for C relative to B, which gave evidence for an error-driven selective attentional learning process.
These studies gave evidence that attentional reallocation occurs in line with the mechanisms of EXIT-like models.
\citeA{inkster2022neural} carried out a direct investigation into brain regions underlying error-driven learning in the IBRE.
Their region of interest (ROI) analysis explicitly targeted areas that were hypothesised to be involved in the computation of prediction error.
They showed that these areas exhibited greater activation during the test phase for $C$ relative to $B$ with a presence of a shared cue during training.
Given these findings, it is reasonable to suggest that prediction-error-driven attentional reallocation occurs in a standard supervised learning paradigm and it is driven by prediction error.

@@

In order to test for the presence of the IBRE, we calculated a Bayes Factor for a one-sample design.
We calculate the probability of responding with the rare label on the critical BC test item, $P(rare|BC)$, for each participant.
Then we tested this distribution of probabilities against the null, $mu = 0.5$, which denoted random responding.
If the Bayes Factor fell below 1/3, we concluded that participants' responses are not different from random responding.
If the Bayes Factor fell above 3, we concluded that participants' responses reliably differ from null.
If the mean probability of $P(rare|BC)$ is higher than 0.5, we conclude that we observed the IBRE.
Values lower than 0.5 would indicate base-rate following.
We used the method implemented in the BayesFactor R package \cite{morey2022bayes}.

@@

\begin{table}[H]
  \begin{center}
    \caption{Group-level mean probabilities for each stimulus presented during the test phase in Experiment 2 after exclusion.\\}
    \label{tab:results-exp2}
    \vskip 0.12in
    \begin{tabular}{rcc}
      \hline
       & $P(common)$ & $P(rare)$ \\
      A & 0.78 & 0.22\\
      AB & 0.95 & 0.05\\
      AC & 0.08 & 0.92\\
      B & 0.92 & 0.08\\
      \textbf{BC} & \textbf{0.27} & \textbf{0.73}\\
      C & 0.07 & 0.93\\
    \end{tabular}
  \end{center}
\end{table}

@@

% LD: so it seems that Johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem because asymmetric representation is hypothesized to be an asymmetry in the way the two categories represented

@@

Another alternative explanation is an eliminative-inference model \cite<ELMO, >{juslin2001high}.
This approach considers the $BC \to rare$ bias to result from rule-learning and post-training inferential processes during $BC$ trials.
The process is most similar to \textit{strategic guessing} \cite{kruschke1995extensions}.
Because of the dissimilarity of $BC$ to the most frequently applied rule, $B \to common$, participants use the most similar rule applicable to $BC$ from a "guessing set", $C \to rare$.
This results in the $BC \to rare$ bias.
Informally, ELMO could accommodate the observational learning paradigm, because of the presence of feature-to-label causality.
Participants could extract the same rules hypothesised to underlie the rare bias in the standard procedure.
Due to this presumed causality, ELMO could hypothesise that people encode rules about symptoms and diseases, which will similarly result in $BC \to rare$ during test.
But it is unclear how it could deal with the cued-recall implementation, as there is no clear-cut feature-to-label relationship in the stimuli presentation that drives rule formation.
% One possibility is to assume that rules are formed between features only, but ELMO is not designed to deal with between-feature relationships.

>>>>

Nonetheless, \citeA{johansen2007paradoxical} demonstrated that the inverse base-rate effect can occur without the traditional predictive learning design.
In one of the conditions in their Experiment 3, the canonical inverse base-rate design (including the shared cue) was implemented in a list format.
In this format, the trial-by-trial presentation of training items was turned into a list of 12 items fitted on a single page.
Subsequently, participants made judgements about new cases on a separate page.
In this condition, participants still exhibited a rare preference on $BC$ trials.
In another condition of Experiment 3, participants received the information about outcome frequencies as a summary before testing.
This summary was presented as prose.
After learning about feature-label information in this manner, participants did not show the IBRE but was matching the base rate.
These experiments give us evidence about another boundary condition for the IBRE - itemised rather than summerised presentation of training items.
% Need to elaborate why this is relevant.

Additionally, there are at least three studies which directly look at error-driven processes in the IBRE.
\citeA{don2019learned}, demonstrated that on $AC$ trials, people fixated on $C$ longer than on $A$ both pre-responding during stimulus presentation and post-responding during feedback (see also \citeA{kruschke2005eye}).
This fixation bias increased with more training.
They also observe greater fixation on $C$ on $AC$ trials, relative to $B$ on $AB$ trials.
Furthermore, \citeA{wills2014attention} in an EEG study observed posterior selection negativity and concurrent frontal positivity for C relative to B, which gave evidence for an error-driven selective attentional learning process.
These studies gave evidence that attentional reallocation occurs in line with the mechanisms of EXIT-like models.
\citeA{inkster2022neural} carried out a direct investigation into brain regions underlying error-driven learning in the IBRE.
Their region of interest (ROI) analysis explicitly targeted areas that were hypothesised to be involved in the computation of prediction error.
They showed that these areas exhibited greater activation during the test phase for $C$ relative to $B$ with a presence of a shared cue during training.
Given these findings, it is reasonable to suggest that prediction-error-driven attentional reallocation occurs in a standard supervised learning paradigm and it is driven by prediction error.

@@

In order to test for the presence of the IBRE, we calculated a Bayes Factor for a one-sample design.
We calculate the probability of responding with the rare label on the critical BC test item, $P(rare|BC)$, for each participant.
Then we tested this distribution of probabilities against the null, $mu = 0.5$, which denoted random responding.
If the Bayes Factor fell below 1/3, we concluded that participants' responses are not different from random responding.
If the Bayes Factor fell above 3, we concluded that participants' responses reliably differ from null.
If the mean probability of $P(rare|BC)$ is higher than 0.5, we conclude that we observed the IBRE.
Values lower than 0.5 would indicate base-rate following.
We used the method implemented in the BayesFactor R package \cite{morey2022bayes}.
\\

@@

\begin{table}[H]
  \begin{center}
    \caption{Group-level mean probabilities for each stimulus presented during the test phase in Experiment 2 after exclusion.\\}
    \label{tab:results-exp2}
    \vskip 0.12in
    \begin{tabular}{rcc}
      \hline
       & $P(common)$ & $P(rare)$ \\
       \hline
      A & 0.78 & 0.22\\
      AB & 0.95 & 0.05\\
      AC & 0.08 & 0.92\\
      B & 0.92 & 0.08\\
      \textbf{BC} & \textbf{0.27} & \textbf{0.73}\\
      C & 0.07 & 0.93\\
    \end{tabular}
  \end{center}
\end{table}

@@

@@

% TODO: mention that ELMO can't do the shared cue effect
Another alternative explanation is an eliminative-inference model \cite<ELMO, >{juslin2001high}.
This approach considers the $BC \to rare$ bias to result from rule-learning and post-training inferential processes during $BC$ trials.
The process is most similar to \textit{strategic guessing} \cite{kruschke1995extensions}.
Because of the dissimilarity of $BC$ to the most frequently applied rule, $B \to common$, participants use the most similar rule applicable to $BC$ from a "guessing set", $C \to rare$.
This results in the $BC \to rare$ bias.
Informally, ELMO could accommodate the observational learning paradigm, because of the presence of feature-to-label causality.
Participants could extract the same rules hypothesised to underlie the rare bias in the standard procedure.
Due to this presumed causality, ELMO could hypothesise that people encode rules about symptoms and diseases, which will similarly result in $BC \to rare$ during test.
But it is unclear how it could deal with the cued-recall implementation, as there is no clear-cut feature-to-label relationship in the stimuli presentation that drives rule formation.
In addition, ELMO also predicts the IBRE in the absence of a shared cue, even though humans do not show the IBRE under those conditions \cite{kruschke2001inverse}.
% One possibility is to assume that rules are formed between features only, but ELMO is not designed to deal with between-feature relationships.
